---
title: Functional Programming in R
fontsize: 9pt
output:
  beamer_presentation:
    theme: "Dresden"
    colortheme: "rose"
    fonttheme: "professionalfonts"
---

```{r echo = F}

set.seed(1)
knitr::opts_chunk$set(collapse = F)

```

# Functions in R

* Functions in are treated just like any other data.
* They are assigned to variables just like any other object. 
* They can be passed to other functions as arguments and returned by other functions as output.
* They can also be anonymous, meaning they can be passed as an argument with out being assigned to a variable.

# `for` loops

You are probably used to `for` loops:

```{r, echo = T}

myData <- c(1, 2, 3, 4, 5)
for(i in 1:length(myData)){
    myData[i] <- myData[i] + 1
}
myData

```
# Functional Programming Basics in R

Alternatives to for loops when working with R objects.

* Vectorized Operations
* Functionals (High Order Functions)

These abstract away the looping construct and pass those instruction's to often highly optimized code that does the looping for the user.

# R objects are immutable (usually)

> Internally, `x$a <- 2` is a combination of two steps: first construct a modified copy of `x`, then change the binding of `x` to the new object; the bindings change, but the original `x` does not. This makes R objects immutable: whenever it looks like you are modifying an object, you are actually creating a modified copy.

> \~ Mutable Objects in R, Hadley Wickham 2010

This means that objects in R can't usually be changed in place. For large objects, this can mean copying big pieces of memory to change small bits of an object.

This is not true for all R objects but its a good rule of thumb to go by unless you know otherwise.

# Vectorized Operations

Vectorized operations simplify common looping tasks, and they are found everywhere in R.

Create a second column that is one more than the first.

```{r}

myData <- 1:15
for(i in 1:length(myData)){
    myData[i] <-  myData[i] + 1
}

```

This is a vectorized version of the same task.

```{r}

myData <- 1:15
myData <- myData + 1

```

# Vector rule of recycling

Vectors of short lengths are recycled. Warnings are thrown when the shorter vector is not a factor of the larger.

```{r}

a <- 1:10
b <- c(2, 10)
a * b

a <- 1:10
b <- 1:3
a * b

```
# Functionals

A function that takes an object and applys a function to each element of that object. 

The function that is applied is passed as an argument. 

In R these are commonly the *apply family of functions. 

* apply - runs on the margins of 2D data
* lapply - returns a list
* sapply - returns a vector, if possible
* mapply - multivariable lapply

# Example

* Assembles a vector of randomly uniform values,
* define a function that will check a value,
* then apply the function to the vector with sapply.

```{r, include = F}

set.seed(1)

```
```{r}

ckFun <- function(x){
    if(x > 50) return(x - 50)
    return(x)
}
x <- round(runif(1000, 1, 100))
print(head(x, 10))
newx <- sapply(x, ckFun)
print(head(newx, 10))

```

# A vectorized approach with `ifelse`

You could also use the `ifelse` function as well. 

This handy function checks if a statement is `TRUE` or `FALSE` then returns a result depending on the condition. 

This would return the same results as the process on the previous slide.

```{r}

xnew <- ifelse(x > 50, x - 50, x)

```

# Benchmarking the different approaches

We will compare the performance of these 3 approaches on a single vector and a column in a data frame.

```{r}

library(microbenchmark)

vecLen <- 1000
vec <- round(runif(vecLen, 1, 100))
dat <- data.frame(col1 = vec)

```

# `for` loops 

```{r}

forV <- function(vec){
    for(i in 1:length(vec)){
        if(vec[i] > 50){
            vec[i] <- vec[i] - 50
        }
    }
    return(vec)
}

forD <- function(dat){
    for(i in 1:nrow(dat)){
        if(dat[i,1] > 50){
            dat[i,2] <- dat[i,1] - 50
        }
    }
    return(dat)
}

```

# `sapply`

```{r}
ckFun <- function(x){
    if(x > 50){
        return(x - 50)
    }
    return(x)
}

funV <- function(vec){
    return(sapply(vec, ckFun))
}

funD <- function(dat){
    dat$col2 <- sapply(dat$col1, ckFun)
    return(dat)
}
```

# `ifelse`

```{r}
ifeV <- function(vec){
    return(ifelse(vec > 50, vec - 50, vec))
}

ifeD <- function(dat){
    dat$col2 <- ifelse(dat$col1 > 50,
                       dat$col1 - 50,
                       dat$col1)
    return(dat)
}
```

# Benchmark Results

```{r, echo = F}

res <- microbenchmark(forV(vec),
                      forD(dat),
                      funV(vec),
                      funD(dat),
                      ifeV(vec),
                      ifeD(dat),
                      times = 10)

boxplot(res)

```

# More about apply functions

`lapply`, `sapply`, and `mapply` are most common.

# `lapply`

`lapply` take a list or a vector as an input, applys a function to each element of the list, then returns a list.

Here, we check each element of the list to see if it is numeric or not.

```{r}

myList <- list(1:5,
               LETTERS[1:10],
               c(T, F, T, T))
lapply(myList, is.numeric)

``` 

# `lapply`

Here, we use an anonymous function that checks if the element is numeric and if so sums the element, else it returns the element.

```{r}

lapply(myList, function(x) if(is.numeric(x)) sum(x) else x)

``` 

# `lapply`

We can use lapply to subset list elements too.

```{r}

lapply(myList, '[', 1)

``` 

# `sapply`

`sapply` simplifies the output of lapply. It'll return a vector instead of a list when it can.

```{r}

sapply(myList, '[', 1)
class(sapply(myList, '[', 1))

```

# `sapply`

```{r}

sapply(myList, '[', 1:3)
class(sapply(myList, '[', 1:3))

```

# `mapply`

`mapply` is multi-variable apply. It can take functions with 2+ inputs and map variables to those inputs.

```{r}

letFun <- function() sample(c("A", "B", "C", "7"), 500, replace = T)
slots <- data.frame(fst = letFun(), snd = letFun(), thd = letFun())
print(head(slots))

```

# `mapply`

```{r}

slotFun <- function(fst, snd, thd){
    if(fst == snd & snd == thd){
        return(paste(fst, snd, thd))
    } else {
        return(NA)
    }
}
runSlots <- mapply(slotFun, slots$fst, slots$snd, slots$thd)
length(which(runSlots == "7 7 7"))

```

# `parallel`

Now that you are thinking functionally, imagine this:

A list of elements.

A single funciton.

Assign a portion of those element to each core and apply the function to those grouped elements.

Re-assemble those elements once they are complete.

```{r, echo = F}

## Resources
## [Functional Progamming - Advanced R](htt://adv-r.had.co.nz/Functional-programming.html)  
## [A Practical Introduction To Functional Programming](https://maryrosecook.com/blog/post/a-practical-introduction-to-functional-programming)  

```
