---
title: 'I/O in R and Variable Binding'
fontsize: 9pt
output:
  beamer_presentation:
    theme: "Dresden"
    colortheme: "rose"
    fonttheme: "professionalfonts"
---

```{r, echo = F}

set.seed(1)

```

# "`<-`" and "`=`"

When assigning a variable in R use "`<-`".

```{r}

x <- 2
y <- 2
x + y

```

# "`<-`" and "`=`"

"`<-`" works both ways, but the reverse usage is uncommon and confusing.

```{r}

2 -> x
2 -> y
x + y

```

# "`<-`" and "`=`"

"`=`" is typically used when assigning arguments in a function.

```{r}

fun <- function(x, y){
    x + y
}
fun(x = 2, y = 2)

```

# Basic I/O

I/O in R varies depending of the data. 

R can connect to nearly any data source you can think of.

```{r}

myData <- data.frame(x = round(runif(1000, 1, 100), 2),
                     y = round(runif(1000, 1, 100), 2),
                     attr = sample(letters, 1000, replace = T))
head(myData, 8)

```

# Basic I/O

`read.csv` and `write.csv` are common and are fine for most tasks.

```{r}

write.csv(myData, "./slides_rmd/data/myData.csv", row.names = F)
myData <- read.csv("./slides_rmd/data/myData.csv")
head(myData, 8)

```

# Spatial Vector Data

```{r, message = F}

library(sf, quietly = T)
myData   <- read.csv("./slides_rmd/data/myData.csv")
myDataSf <- st_as_sf(myData, coords = c("x", "y"))
head(myDataSf)

```

# Spatial Vector Data

```{r}

plot(myDataSf)

```

# Spatial Vector Data

```{r}

write_sf(myDataSf,
         "./slides_rmd/data/myDataSf.shp",
         driver = "ESRI Shapefile")
myDataSf <- read_sf("./slides_rmd/data/myDataSf.shp")

```

# Spatial Vector Data

```{r}

plot(myDataSf)

```

# Spatial Vector Data

```{r}

plot(myDataSf)

```

# Spatial Raster Data

```{r, messages = F}

library(raster, quietly = T)
myMatrix <- matrix(sort(round(runif(10000, 1, 100))), nrow = 100)
myRaster <- raster(myMatrix)

```

# Spatial Raster Data

```{r}

plot(myRaster)

```

# Spatial Raster Data

```{r}

writeRaster(myRaster,
            "./slides_rmd/data/myRaster.tif",
            overwrite = T)
myRaster <- raster("./slides_rmd/data/myRaster.tif")

```

# Spatial Raster Data

```{r}

plot(myRaster)

```

# Saving R Objects

`saveRDS()` and `readRDS()` are the preferred methods for saving R objects to disk when interoperability is not important.

```{r}

myRaster <- raster("./slides_rmd/data/myRaster.tif")
saveRDS(myRaster, "./slides_rmd/data/myRaster.rds")

```

# Saving R Objects

```{r}

myRaster <- readRDS("./slides_rmd/data/myRaster.rds")
class(myRaster)

```
